---
title: "Shell scripts for independent contrast regressions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shell scripts for independent contrast regressions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The main goal of the __InNOutBT__ package is to readily create input data to and
process output data from Meade & Pagel's (2022) [BayesTraits](http://www.evolution.reading.ac.uk/BayesTraitsV4.0.0/BayesTraitsV4.0.0.html) 
computer package.\

Here in this article, we show how to use the package's function `BayesTraits.shell`
for producing shell scripts and associated run commands of BayesTraits phylogenetic 
independent contrast regression analyses.


# Setup

__InNOutBT__ is not on CRAN yet but you can install the latest development version 
from [GitHub](https://github.com/):

``` r
#install.packages("devtools")
devtools::install_github("DBOSlab/InNOutBT")
```
```{r, eval = FALSE}
library(InNOutBT)
```


# Create sample of trait data using a different subset of explanatory variables

This function produces all BayesTraits input data of linked samples of trait/climate 
data (distribution data), by automatically generating two .txt files: one including 
the linked traits and the other with mean values for each included trait. The 
function also includes an argument to calculate the net node count for each
species/accession given a phylogenetic tree of reference, which has been successfully 
used to explore the effect of speciation rates (e.g. [O’Donovan et al. 2018](https://doi.org/10.1038/s41559-017-0454-6)) on the response variable.\

The example below shows how to mine all accepted species for a specified vector 
of families. 

```{r, eval = FALSE}
library(InNOutBT)

BayesTraits.shell(meanfile_dir = "results_BayesTraits_input/03Jun2022",
                  linkfile_dir = "results_BayesTraits_input/03Jun2022",
                  treefile_dir = "Data",
                  BayesTraits_dir = "/Users/domingoscardoso/BayesTraitsV4",
                  responvar = "DBH",
                  treetransf = c("UNI", "Lambda", "VR"),
                  bi = "30000000",
                  it = "130000000",
                  sa = "100000",
                  st = c("500", "50000"),
                  syst = "unix",
                  dir_create = "results_BayesTraits_shell",
                  cc_DataTree = TRUE)
```


# Create sample of trait data using a different subset of explanatory variables

To mine a species checklist for all families of flowering plants, you recommend to
load the dataframe-formatted data object called `POWOcodes` that comes associated with
the __expowo__ package. Because the `POWOcodes` data object already contains the 
URI addresses for all angiosperms families recognized in the [POWO](https://powo.science.kew.org) 
database, you do not need to perform the time-consuming URI search using __taxize__. \

The example below shows how to mine a global checklist of all accepted species of
flowering plants by using the vector of all angiosperm families and associated URI 
addresses stored in the `POWOcodes` object.

```{r, eval = FALSE}
BayesTraits.shell(meanfile_dir = "results_BayesTraits_input/03Jun2022",
                  linkfile_dir = "results_BayesTraits_input/03Jun2022",
                  treefile_dir = "Data",
                  BayesTraits_dir = "/Users/domingoscardoso/BayesTraitsV4",
                  responvar = "DBH",
                  treetransf = c("UNI", "Lambda", "VR"),
                  bi = "30000000",
                  it = "130000000",
                  sa = "100000",
                  st = c("500", "50000"),
                  syst = "unix",
                  dir_create = "results_BayesTraits_shell",
                  cc_DataTree = TRUE)
```
